---
title: "Using spatial model to estimate gene parameters"
output: html_notebook
---


```{r}
library(SFEData)
library(SingleCellExperiment)
library(SpatialFeatureExperiment)
library(Voyager)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```
Load Visium MSK dataset from `SFEData`
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.
```{r}
sfe <- McKellarMuscleData("small")
load("../data/vis_slim_v1.RData") # Contains author cell type labels
```
```{r}
# Subset a sfe object
sfe_tissue <- sfe[,colData(sfe)$in_tissue]
colGraph(sfe_tissue, "visium") <- findVisiumGraph(sfe_tissue)

features_use <- names(rowSums(counts(sfe_tissue)) > 10)
```
# Run Moran's I
Here, I'm only using the spots that overlap tissue to estimate Moran's I. Because most empty spots do not have RNA, there is a sort of "built-in" spatial correlation between tissue containing and empty spots. This is not the relationship that we're interested in, however. In the cells below, I will compute Moran's I on a subset of the features and focus on genes with high spatial autocorrelation. 
```{r}
sfe_tissue <- runMoransI(sfe_tissue,
    features = features_use, colGraphName = "visium",
    exprs_values = "counts"
  )
```
Take a look at some of the results
```{r}
moran_order <- order(rowData(sfe_tissue)[["moran_Vis5A"]], decreasing = TRUE, na.last=NA)
moran_res <- rowData(sfe_tissue)[moran_order, c('Ensembl', "symbol", "moran_Vis5A")]

head(moran_res, 10)
```
Plot the in/out tissue metadata
```{r}
plotSpatialFeature(sfe, c("in_tissue",'nCounts'),
    colGeometryName='spotPoly', annotGeometryName = 'myofiber_simplified')
```

Want to locate genes that have expression in empty spots, to fit the Poisson model. Recall, genes are on the rows
```{r}
tissue <- colData(sfe)$in_tissue
n_exprs_empty <- rowSums(SingleCellExperiment::counts(sfe[,!tissue]) > 0)
genes_exprs_empty <- rownames(sfe)[which(n_exprs_empty >= 10)]

gene_mask <- rowData(sfe)$Ensembl %in% genes_exprs_empty

# rowSums(counts(sfe[gene_mask, !tissue]))
```
Barcode matching to obtain cell identities 
```{r}
bcs <- colData(sfe)$barcode
author_idents <- Seurat::Idents(vis.list[[1]])
names(author_idents) <- stringr::str_replace(names(author_idents), "-[0-9]+", "")

ct_mask <-  which(names(author_idents) %in% bcs)

ct <- rep(NA, ncol(sfe))
ct[bcs %in% names(author_idents)] <- (author_idents[ct_mask])

colData(sfe)$author_idents <- ct
colData(sfe)[!tissue, "author_idents"] <- "empty"
```
Plot "cell type"  label
```{r}
plotSpatialFeature(sfe, c("author_idents", "in_tissue"),
    colGeometryName='spotPoly', annotGeometryName = 'myofiber_simplified')
```

Plot spatial features
```{r}
plotSpatialFeature(sfe, c("ENSMUSG00000025172", "ENSMUSG00000030470", "ENSMUSG00000056328"),
    exprs_values='counts', annotGeometryName = "myofiber_simplified")
```

Compute relevant spatial graphs...
```{r}
# full adjacency matrix
g <- findVisiumGraph(sfe, style="B")
W <- spdep::nb2mat(g$neighbours, style="B")

# For empty spots...
g <- findVisiumGraph(sfe[,!tissue], style="B")
W_ext <-  spdep::nb2mat(g$neighbours, style="B")

# For tissue spots...
g <- findVisiumGraph(sfe_tissue, style="B")
W_int <-  spdep::nb2mat(g$neighbours, style="B")
```
Compute estimated means for poisson model

```{r, pois-border}
pois.file <- fs::path("../inst/stan/pois", ext="stan")

# Compile and run
pois.mod <- cmdstanr::cmdstan_model(pois.file)
```

Set up data for fitting

```{r}
genes_exprs_empty <- which(features_use %in% genes_exprs_empty)
genes_exprs_empty <- features_use[genes_exprs_empty]

# nodes <- which(W_ext==1, arr.ind=TRUE)
# nodes <- nodes[nodes[,1] < nodes[,2],]

data <- list(
  N = dim(W_ext)[1], 
  count_n = sum(!tissue), 
  #count_inds = which(!tissue),
  count_inds = seq(sum(!tissue)),
  W = as.matrix(W_ext)
)

```
Finally, the fit for all genes
```{r}
pois.mods <- lapply(genes_exprs_empty, function(i){

  data$mu <-  rep(mean(counts(sfe)[i,!tissue])/0.5, sum(!tissue))
  data$counts <- counts(sfe)[i,!tissue]
  
  
  mod <-  pois.mod$optimize(data)
  mod
})
```

```{r}
mods_params <- purrr::map(pois.mods, \(x) x$summary(c("car_rho", "car_rho_trans", "car_scale")))
mods_params <- bind_rows(mods_params, .id = 'group')

mods_params <- mods_params |>
  pivot_wider(
    id_cols = group,
    names_from = variable,
    values_from = estimate) 

hyperpars <- summarise(mods_params, across(where(is.numeric), mean))
```

Now, we will fit the interior spots 
```{r stan_setup}
file <- fs::path("../inst/stan/car",ext="stan")
mod <- cmdstanr::cmdstan_model(file)

```

Data set up...
```{r}
g <- 'ENSMUSG00000056328'
keep <- colData(sfe)[tissue, "author_idents"] %in% c(4, NA)

y <- counts(sfe)[g, tissue][keep]

nodes <- which(W_int==1, arr.ind=TRUE)
nodes <- nodes[nodes[,1] < nodes[,2],]

data <- 
  list(
    N = sum(tissue),
    #car_scale = hyperpars[['car_scale']], 
    car_scale=1, 
    #car_rho =   hyperpars[['car_rho']], 
    car_rho = 0.33,
    W = W_int,
    counts = y, 
    count_n = sum(keep),
    count_inds = which(keep),
    #count_inds = seq(sum(tissue)),

    bs_mu = 0, 
    bs_var = 2, 
    mu_mu = 0, 
    mu_var = 1
  )
```
Fit the NB ICAR model
```{r}
# icar_mod <- cmdstanr::cmdstan_model("./inst/stan/nb_icar.stan")
```

Now the fit
```{r}
# fit <- mod$sample(data, chains=2, iter_warmup = 1000, show_messages=FALSE)
fit <- mod$optimize(data)
fit$summary(c("trans_rate", "beta_"))
# fit$summary(paste0('rates[',1:57,']'))$estimate

trans_rate <- fit$summary("trans_rate")$estimate
bs <- fit$summary("beta_")$estimate
mean_est <- mean(fit$summary(paste0('rates[',1:55,']'))$estimate)

trans_rate / bs / mean_est
``` 

Let's try to fit for a sample of the genes from the dataset
```{r}
genes <- sample(genes_exprs_empty, 20, replace=FALSE)
fits_all_20 <- purrr::map(genes, \(g){
  data$counts <-  counts(sfe)[g,tissue]
  data$count_n <- length(counts(sfe)[g,tissue])
  data$count_inds = seq(sum(tissue))
  
   
  mod$sample(data, chains=2, show_messages=FALSE)
})

```
Wrangle fits
```{r}
names(fits_all_20) <- genes
fits_all_20 <- lapply(fits_all_20, 
                      function(x) x$summary(
                        c('mu', 'log_bs', "burst_size", paste0('rates[',1:57,']')))
                      ) %>%
  bind_rows(.id='gene') 

fits_all_20 <-  fits_all_20 %>% 
  select(gene, variable, mean) %>% 
  pivot_wider(names_from='variable', values_from = 'mean')
```
observed gene means
```{r}
obs <- data.frame(gene = genes, obs_mean2 = rowMeans(counts(sfe)[genes,tissue]))

fits_all_20 <- left_join(fits_all_20, obs)
```
Plot some results
```{r}
# Mean fit with observed mean? 
ggplot(fits_all_20, aes(mu, obs_mean2, label=gene)) + 
  geom_point(
    shape=21,
    stroke=1
  ) + 
  #geom_label() + 
  geom_abline(slope=1, intercept=0) +
  theme_bw()
```
Plot some histograms
```{r}
ps <- map(genes, \(x){
  p <- data.frame(gene =x, counts=counts(sfe)[g,tissue])%>%
    ggplot(aes(x = counts)) + 
    geom_histogram()
  
  mu <- fits_all_20 %>% filter(gene %in% x) %>% pull(mu)
  logbs <- fits_all_20 %>% filter(gene %in% x) %>% pull(log_bs)
  bs <- 10^logbs
  
  vals <- rnbinom(55, size=mu, prob=bs/(1+bs))
  # p2 <- data.frame(sampled = vals) %>%
  #   ggplot(aes(x = sampled)) + 
  #   geom_density(colour = 'red')
  
  p + 
    geom_histogram(
      aes(x = sampled), 
      data =  data.frame(sampled = vals),
      color = 'red', alpha=0.4) +
    theme_bw()
})

```
Display
```{r}
#Display 
cowplot::plot_grid(plotlist=ps, nrow = 4)
```

