---
title: "Modeling Spatial Technical Varition"
output: html_notebook
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/kaylajackson/Library/CloudStorage/OneDrive-CaliforniaInstituteofTechnology/Projects/spatial_normalization/normalize')
```

# Loading Modules
```{r mods}
library(Matrix)
library(sf)
library(spdep)
library(cmdstanr)
library(StanHeaders)

library(fs)
library(readr)
library(magrittr)

library(glue)
library(stringr)

library(tibble)
library(purrr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(latex2exp)

library(posterior)
library(bayesplot)

set.seed(3422)
```

# Define helper functions
```{r funcs}
# Find rook style neighbors
st_rook = function(a, b = a) st_relate(a, b, pattern = "F***1****")

# Convert sgbp to nb
as.nb.sgbp <- function(x, ...) {
  attrs <- attributes(x)
  x <- lapply(x, function(i) { if(length(i) == 0L) 0L else i } )
  attributes(x) <- attrs
  class(x) <- "nb"
  x
}

# Check if ranges overlap
check_overlap <- function(data, x, y){
  x = data[[x]]
  y = data[[y]]
  
  if (length(x) != length(y)) return("length problems")
  
  grid <- tidyr::expand_grid(x1=seq_along(x), x2=seq_along(y)) 
  grid <- dplyr::filter(grid, x1 < x2)
  
  
  out <- purrr::pmap_dfr(grid, \(x1,x2){
    c(id1=x1, id2=x2, res = `::`("DescTools", `%overlaps%`)(c(x[x1], y[x1]), c(x[x2],y[x2])))
  })
  
  key <- dplyr::filter(out, res==1) %>% 
    dplyr::mutate(key=paste0(id1,id2)) %>% 
    dplyr::pull(key)
  
  purrr::map(seq_along(x), \(ind){
    finds <- stringr::str_detect(key, as.character(ind))
    stringr::str_remove(key[finds], as.character(ind))
  })
}
```
# Load Data
```{r}
# Load physical parameters
burst_params <- read_csv("../data/Larsson_c57_kinetic_params.csv", skip=1,
                         col_names = c("gene", "mean_exprs","bs", 'bs_conf'),
                         show_col_types = FALSE)

burst_params <-
  burst_params %>%
  filter(mean_exprs > 1) %>%
  mutate(
    bs_conf = str_remove_all(bs_conf, "\\[|\\]"),
    bs_conf = str_squish(bs_conf)
  ) %>%
  separate_wider_delim(
    bs_conf, delim=" ", 
    names = c(NA, "bs_conf_ll", "bs_conf_ul"),
    cols_remove=FALSE) %>%
  type_convert() %>%
  filter(!is.na(bs_conf_ll))

```

# Define constants and hyperparameters
```{r}
# Parameters for cell/gene simulation
grid_size = 10
n_celltypes = 2
n_genes = 100

labels <- letters[seq(n_celltypes)]
remainder <- grid_size^2 %% n_celltypes

type_size <- rep(floor(grid_size^2/n_celltypes), n_celltypes)
type_size[n_celltypes] <- type_size[n_celltypes] + remainder
names(type_size) <- labels

# How will physical kinetic parameters be sampled?
resample <- nrow(burst_params) < n_genes

# Parameters affecting structure of technical variance
rho <- 0.99
sigma_sq <- 1
```

# Generate neighborhood graph 

```{r}
sfc = st_sfc(st_point(c(0,0)), st_point(c(grid_size,grid_size)))
grd =  st_make_grid(sfc, n = c(grid_size,grid_size))

sfc_sgbp_rook = st_rook(grd)

sf_nb_rook = as.nb.sgbp(sfc_sgbp_rook)
col.w <- nb2listw(sf_nb_rook)


# Style = B is the binary coding. 
# Style = W is the row normalized. 
W <- nb2mat(sf_nb_rook, style="B")
nn <- rowSums(W)

borders <- which(nn < 4)
```

```{r}
dfc <- st_sf(geometry = grd)
```

## Separate neighborhood graph for interior points vs. border points
```{r}
# Exterior
nb_rook_ext = st_rook(dfc[borders,])
nb_rook_ext = as.nb.sgbp(nb_rook_ext)
col.w.ext <- nb2listw(nb_rook_ext)

W_ext <- nb2mat(nb_rook_ext, style="B")
D_ext <- Diagonal(length(borders), rowSums(W_ext))

# Interior
nb_rook_int = st_rook(dfc[-borders,])
nb_rook_int = as.nb.sgbp(nb_rook_int)
col.w.int <- nb2listw(nb_rook_int)

W_int <- nb2mat(nb_rook_int, style="B")
D_int <- Diagonal(length(borders), rowSums(W_int))
```

# Generate covariance matrix 
Here, the covariance matrix  encodes how technical variance is structured across the lattice
```{r}
D <- diag(100)
diag(D) <- nn

IW = D - rho*W
OMEGA <- (1/sigma_sq) * IW
```

# Construct technical variance
Use the copula function to construct a multivariate distribution with beta marginals.
Each marginal distribution has the following form: 

$X_i \sim \text{Beta}(0, 10)$

### Change of idea
Now, 

$\test{logit}(p) \sim \text{MVN}(0, Q^{-1})$, where

$Q^{-1}$ is the precision matrix constructed from the lattice and neighborhood matrix. In this construction we guarantee that $Q^{-1}$ is non-singular, and thus, we have a proper CAR model to model spatial correlation.  

```{r}
# Sample from MVN parameterized by precision matrix
# Samples are drawn using Cholesky and STD MVN

Z <- stanFunction("multi_normal_prec_rng", mu=rep(0,100), S=OMEGA)
# CDF_Z <- pnorm(Z, 0, diag(solve(OMEGA)))
# theta  <- qbeta(CDF_Z, 1, 10)

theta <- 1/(1+exp(-Z))
```

## Confirm that technical variance is spatially correlated
```{r}
(moran(theta, col.w, n=length(sf_nb_rook), Szero(col.w)))
```

```{r}
dfc$theta <- theta
plot(dfc["theta"])
```

# Simulate Burst Parameters for "genes"
```{r}
dfc$type <- rep(labels, type_size)

dfc <- dfc %>% group_by(type) %>% 
  mutate(cellid = paste(type, row_number(), sep="_")) %>% 
  ungroup() %>%
  mutate(border = ifelse(row_number() %in% borders, TRUE, FALSE))

dfc <- mutate(dfc, type_border = glue("{type}_{border}"))

# type_size_new <- dfc %>% group_by(type_border) %>% tally() %>% pull(n, type_border)
# type_size_new <- names(type_size_new)

burst_params_int <- lapply(seq(n_celltypes), function(x){
  rows = sample(nrow(burst_params), n_genes, replace=resample)
  burst_params[rows,] %>% 
    mutate(
      type = labels[x],
      type_border = glue("{type}_FALSE"),
      cellid = list(dfc$cellid[dfc$type_border == glue("{type}_FALSE")]),
      p = list(dfc$theta[dfc$type_border == glue("{type}_FALSE")]),
      gene = paste0("gene_", row_number()))
})

burst_params_int <- bind_rows(burst_params_int) %>% arrange(gene, type)
burst_params_int$bs_p <- pmap(burst_params_int, \(bs, p, ...) bs*p)

# Update mean expression around the border of the grid
burst_params_ext <- burst_params_int %>%
  split(.$gene) %>%
  map(\(df) {
    means <- rep(df$mean_exprs, type_size)
    means[borders] <- 0

    b_means1 <- crossprod(W, means)[,1]/rowSums(W)
    corners <- which(b_means1 == 0)

    b_means2 <- crossprod(W, b_means1)[,1]/rowSums(W)
    b_means1[corners] <- b_means2[corners]

    data.frame(mean_exprs = b_means1[borders], p = dfc$theta[borders],
               type_border = dfc$type_border[borders],
               cellid = dfc$cellid[borders])
    })

burst_params_ext <- bind_rows(burst_params_ext, .id = "gene")
```

Finally, sample counts with parameters
```{r}
# sample counts with technical variation
counts_int <- burst_params_int %>% 
  mutate(row = row_number()) %>% 
  group_by(row) %>% 
  group_map(~  rnbinom(length(.x$bs_p[[1]]), size=.x$bs_p[[1]], mu=.x$mean_exprs))

burst_params_int$counts <- counts_int

counts_int <- burst_params_int %>% 
  unnest(c(counts, cellid)) %>% 
  pivot_wider(id_cols=gene, names_from=cellid, values_from=counts)

burst_params_ext <- burst_params_ext %>% 
  rowwise() %>% 
  mutate(counts = rpois(1, mean_exprs*p))
  
counts_ext <- burst_params_ext %>% 
  pivot_wider(id_cols=gene, names_from=cellid, values_from=counts)

counts <- merge(counts_int, counts_ext, by = 'gene')
counts <- column_to_rownames(counts, var = "gene")

ij_inds <- which(counts != 0, arr.ind = TRUE)
x_inds <- apply(ij_inds, 1, \(x) counts[[x['row'],x['col']]])

mtx <- sparseMatrix(i=ij_inds[,"row"], j=ij_inds[,"col"], x=x_inds,
                    dimnames=list(rownames(counts), colnames(counts)))

counts <- counts %>%
  rownames_to_column('gene') %>% 
  pivot_longer(-gene, names_to='cellid') %>% 
  pivot_wider(names_from ='gene')
```

Add counts to sf dataframe
```{r}
dfc <- merge(dfc, counts, by = 'cellid')
ord <- str_sort(dfc$cellid, numeric = TRUE)
dfc <- dfc[match(ord, dfc$cellid),]
```

Add Moran's I coefficient to gene metadata. Interestingly, even though burst size is correlated across the grid, this does not guarantee that counts are correlated. What is says is that you can observe a gene that does not have spatial correlation, even though technical variation is spatially correlated. 
```{r}
res_moran <- burst_params_int %>%
  group_by(gene) %>%
  group_map(~  moran(unlist(.x$counts), col.w.int, n=length(nb_rook_int), Szero(col.w.int))$I)

res_moran <- data.frame(moran = unlist(res_moran), gene = unique(burst_params_int$gene))
burst_params_int <- merge(burst_params_int, res_moran, by = "gene")
```

Are there any genes that might be of interest ?
```{r}
# genes <- sim_burst_params %>% dplyr::filter(!str_detect(overlaps, "1|2|3|4")) %>% 
#   distinct(gene) %>% pull(gene)
genes <- burst_params_int %>% distinct(gene, .keep_all=TRUE) %>% 
  slice_max(abs(moran), n=20) %>% pull(gene)

plot(dfc[genes], max.plot=20)
```
# Fit Poisson CAR model with Stan
```{r, pois-border}
pois.file <- path("../inst/stan/pois", ext="stan")

# Compile and run
pois.mod <- cmdstan_model(pois.file)
```
Data set up for border model
```{r}
g <- 'gene_91'

nodes <- which(W_ext==1, arr.ind=TRUE)
nodes <- nodes[nodes[,1] < nodes[,2],]

# Set up data for STAN
y <- counts_ext %>% filter(gene %in% g) %>% select(-gene) %>% unlist()

# MU computation
mns <- dfc[[g]]
mns[borders] <- 0

mu <- crossprod(W, mns)
tmp <- crossprod(W, mu)
corners <- which(nn == 2)
mu[corners] <- tmp[corners]

mu <- c(mu)/rowSums(W)
mu <- mu[borders]

data <- list(
  N = length(borders), 
  N_edges = nrow(nodes), 
  D=as.matrix(D_ext),
  W = as.matrix(W_ext),
  mu = mu,
  #mu = rep(1, 36),
  counts = y
)

```

```{r}
# errors if mu = 0
fit.pois <- pois.mod$sample(data, chains=2)
```

```{r}
draws <- as_draws_matrix(fit.pois)
yrep <- draws[, glue::glue('counts_ppc[{1:36}]')]

ppc_dens_overlay(y, yrep[1:200, ])
```

Some other diagnositics? Did we do an ok job? See `blah.R`
# Fit NB model with Stan
The model to fit is as follows

$Y_{cg} \sim \text{Neg.Binomial}(\frac{\alpha}{\gamma}, bp)$

where, $Y_{cg}$ represents the observed count in cell $c$ of gene $g$. The parameters $\alpha, \gamma,$ and $b$ are gene specific transcription rates, degradation rates, and burst sizes, respectively. The paramter $p$ is a technical parameter that represents the probability that a molecule is captured at each grid location. As written, the parameters are not uniquely identifiable. 

```{r stan_setup}
file <- path("../inst/stan/car",ext="stan")
mod <- cmdstan_model(file)
# mod$print()
```

## Format data for stan
```{r}
# Test on one gene
g <- 'gene_52'

y <- dfc[dfc$type_border == "b_FALSE",][[g]]
hyperpars <- fit.pois$summary(c("car_rho", "car_scale"))$mean

# cndl_means <- crossprod(W, y)[,]/nn
# log_cndl_means <-  log(cndl_means+.Machine$double.eps,10)

data <- 
  list(
    N = grid_size^2,
    W = W,
    count_n = length(y),
    count_inds = which(dfc$type_border == 'a_FALSE'),
    car_rho=hyperpars[1],
    car_scale=hyperpars[2],
    bs_mu = 0,
    bs_var = 1, 
    mu_mu = 0, 
    mu_var = 5, 
    counts = y
  )
```

```{r}
burst_params_int %>% 
  filter(gene%in%g) %>% 
  select(gene, type, mean_exprs, bs) %>% 
  mutate(logbs = log(bs, 10))
```

```{r}
fit <- mod$sample(data, chains=2, num_warmup = 2000, show_messages=FALSE)
fit$summary(c("mu", "log_bs"))
```

```{r}
data <- 
  list(
    N = grid_size^2,
    W = W,
    car_rho=hyperpars[1],
    car_scale=hyperpars[2],
    bs_mu = 0,
    bs_var = 1, 
    mu_mu = 0, 
    mu_var = 5 
    # counts = y
  )

fits_all <- burst_params_int %>%
  group_by(gene, type) %>%
  group_map(~ {
    data$counts <- .x$counts[[1]]
    data$count_n <- length(.x$counts[[1]])
    data$count_inds <- which(dfc$type_border == .x$type_border[1])
    
    mod$sample(data, chains=2, show_messages=FALSE)$summary(c("mu", "log_bs"))
    
  })
```

Wrangle the fits
```{r}
nm <-  glue("{burst_params_int$gene}_{burst_params_int$type}")
names(fits_all) <- nm

fits_all <- bind_rows(fits_all, .id="id") %>% 
  mutate(type = str_split_i(id, "_", -1), gene = str_remove(id, "_[ab]")) %>% 
  select(-id)

fits_all_mean <- fits_all %>% select(type, gene, variable, mean) %>% 
  pivot_wider(names_from = variable, values_from = mean)

```
Add burst parameters to fits
```{r}
fits_all_mean2 <- fits_all_mean %>%
    merge(burst_params_int)  %>% 
    select(type, gene, mu, log_bs, mean_exprs, bs, p)

fits_all_mean2$phi_nb <- pmap(fits_all_mean2, \(log_bs, p,...) (10^log_bs)*p)

# Compute variance at each spatial location
fits_all_mean2$var_nb <- pmap(fits_all_mean2, \(mu, phi_nb,...) mu * (1 + mu/phi_nb))
```
# Gene stats
```{r}
# gene_stats <- fits_all_mean2 %>% 
#   group_by(gene) %>% 
#   group_map(\(df, name) c(name, evpv = mean(unlist(df$var_nb)/2))) %>% 
#   bind_rows()
# gene_stats$var_nb <- rowvars[gene_stats$gene]
data_p1 <- fits_all_mean %>% 
  mutate(bs = 10^log_bs, var = mu * (1 + mu/bs)) %>% 
  group_by(gene) %>% summarise(unexpl_ct_var = mean(var), expl_ct_var = var(mu))  %>% 
  mutate(var_gene = unexpl_ct_var + expl_ct_var, frac_unexpl = unexpl_ct_var/ var_gene) %>%
  merge(distinct(burst_params_int[c('gene', 'moran')]))


p1 <-  ggplot(data_p1, aes(frac_unexpl, moran)) +
  geom_vline(
    xintercept=0.50,
    linetype='dashed',
    alpha=0.75,
    color='red') + 
    geom_hline(
      yintercept = 0,
      linetype='dashed',
      alpha=0.75,
      color='red')+ 
    geom_point(shape=21, fill='gray', alpha=0.8,size=rel(2))  +
  theme_bw() + 
  theme(
    aspect.ratio=1,
    panel.grid = element_blank(),
  )

ggsave('moran_fracunexpl_11142023.png', p1, width=2,height=2, scale=2)
```

```{r}
p1 <- fits_all_mean %>%
  merge(burst_params_int) %>%
  ggplot(aes(y=log(bs,10), x=log_bs)) + 
  geom_abline(
    slope=1,
    intercept=0,
    linewidth=1.5,
    linetype='dashed',
    alpha=0.75,
    color='red'
  ) + 
  geom_point(shape=21, fill='gray', alpha=0.8,size=rel(2)) +
  theme_bw() + 
  labs(
    x = TeX(r'(log($\hat{b}$))'),
    y = TeX(r'(log($b$))')
  )+
  theme(
    aspect.ratio=1,
    panel.grid = element_blank(),
  )

ggsave('burst_size.png', p1, width=2,height=2, scale=2)
```

```{r}
# a simple function that takes in vector of counts and stan model, and returns estimated parameters
stan.optim <- function(model, data, counts,...){
  data$counts <- counts
  
  fit <- model$optimize(data, ...)
  fit$summary()
  
  }
```
## Fit the model
Note, each cell type should be fit separately 
```{r}
model_fits <- sim_burst_params %>% 
  split(.$gene) %>%
  map(\(x){
    stan.optim(mod, data, x$counts[[1]]) %>%
      pivot_wider(names_from="variable", values_from='estimate', 
                  names_glue='fit_{variable}') 
  })

model_fits <- list_rbind(model_fits, names_to = 'gene')

model_fits <- merge(sim_burst_params, model_fits, by = 'gene') %>%
  select(gene, fit_mu, mean_exprs, fit_burst_size, bs, type)

```

```{r}
ggplot(model_fits, aes(y=log(bs), x=log(fit_burst_size))) + 
  geom_point() + 
  geom_abline(slope=1, intercept=0, color='red')
```

```{r}
ggplot(model_fits, aes(y=log(mean_exprs), x=log(fit_mu))) + 
  geom_point() + 
  geom_abline(slope=1, intercept=0, color='red')
```


# Some debugging
```{r}
# Based on these results, I think somehow the counts and parameters are mixed up
out <- map_df(seq(10), \(x){
  y <- rnbinom(100, size=bs*theta, mu=mm)
  stan.optim(mod, data, y, theta)
}, .id = 'iter')
```


# Lab meeting results
```{r}
# a simple function that takes in vector of counts and stan model, and returns estimated parameters
stan.optim <- function(model, data, counts, rates, ...){
  data$counts <- counts
  data$count_n <- length(counts)
  data$rates <- rates
  
  fit <- model$optimize(data, ...)
  fit$summary()
  
  }
```

```{r}
file <- path("../inst/stan/car_prob",ext="stan")
mod <- cmdstan_model(file)
```

```{r}
data <- 
  list(
    bs_mu = 0,
    bs_var = 5, 
    mu_mu = 0, 
    mu_var = 10 
  )

model_fits <- burst_params_int %>%
  #filter(gene == 'gene_1') %>%
  group_by(gene, type) %>%
  group_map(\(df, name){
    stan.optim(mod, data, df$counts[[1]], df$p[[1]]) %>%
      pivot_wider(names_from="variable", values_from='estimate',
                  names_glue='fit_{variable}')
  })
# gs <- glue("gene_{sample(100, 20)}")
# model_fits <- burst_params_int %>%
#   filter(gene %in% gs) %>%
#   group_by(gene, type) %>%
#   group_map(\(df, name){
#     
#     data$counts <- df$counts[[1]]
#     data$count_n <- length(df$counts[[1]])
#     data$rates <- df$p[[1]]
#     mod$sample(data, chains=2, show_messages=FALSE)$summary(c("mu", "log_bs"))
#   })

names(model_fits) <- glue("{burst_params_int$gene}_{burst_params_int$type}")
model_fits <- list_rbind(model_fits, names_to = 'gene')

model_fits <- model_fits %>%
  mutate(type = str_remove(gene, "gene_[0-9]+_"),
         gene = str_remove(gene, "_[ab]")
         ) %>%
   # pivot_wider(id_cols = c("gene", "type"), 
   #             names_from=variable, 
   #             values_from = 'mean', 
   #             names_glue = "fit_{variable}") %>% 
  merge(burst_params_int, by = c("gene", "type"), all.x = TRUE) %>%
  select(gene, type, fit_mu, mean_exprs, bs, fit_log_bs)

model_fits <- model_fits %>%
  mutate(log_bs = log(bs, 10))
# model_fits <- merge(sim_burst_params, model_fits, by = 'gene') %>%
#   select(gene, fit_mu, mean_exprs, fit_burst_size, bs, type)
```

```{r}
p8 <- model_fits %>% 
  mutate(diff_log = abs(fit_log_bs-log_bs)) %>% 
  filter(diff_log < .3) %>%
  ggplot(aes(log(fit_mu,10), log(mean_exprs,10))) + 
  geom_abline(
    slope=1,
    intercept=0,
    linewidth=1.5,
    linetype='dashed',
    alpha=0.75,
    color='red'
  ) + 
  geom_point(shape=21, fill='gray', alpha=0.8,size=rel(2)) +
  theme_bw() + 
  labs(
    x = TeX(r'(Fitted log($\hat{k/\gamma}$))'),
    y = TeX(r'(True log($k/\gamma$))')
  )+
  theme(
    aspect.ratio=1,
    panel.grid = element_blank(),
  ) 

p8
ggsave('mu_givep.png', p8, width=2,height=2, scale=2)
```